---
title: "RR2 first draft"
format: 
  html:
    self_contained: true
    toc: true
    embed-resources: true
editor: visual
author: Chita Chaaban
date: Winter 2026
---

## First Draft: 

### Introduction

### w2eq15

How likely unable to pay for household necessities next 12 months

The next questions ask you to think about what might happen during the next 12 months.

In the next 12 months, how likely is it that there will be periods when you are unable to pay for your basic household necessities?

| Value | Category          |
|:------|:------------------|
| 1     | Not at all likely |
| 2     | Not very likely   |
| 3     | Somewhat likely   |
| 4     | Very likely       |
| 5     | Extremely likely  |
| 9     | No answer\*       |

\*) Missing value

### w2eq16

How likely unable to afford a one-week holiday away from home next 12 months

In the next 12 months, how likely is it that you will be unable to afford a one-week holiday away from home?

| Value | Category          |
|:------|:------------------|
| 1     | Not at all likely |
| 2     | Not very likely   |
| 3     | Somewhat likely   |
| 4     | Very likely       |
| 5     | Extremely likely  |
| 9     | No answer\*       |

\*) Missing value

### w2eq17

How likely unable to pay for unexpected expenses next 12 months

In the next 12 months, how likely is it that you will be unable to pay for unexpected expenses?

| Value | Category          |
|:------|:------------------|
| 1     | Not at all likely |
| 2     | Not very likely   |
| 3     | Somewhat likely   |
| 4     | Very likely       |
| 5     | Extremely likely  |
| 9     | No answer\*       |

\*) Missing value

### w2eq18

How likely unemployed and not able to find job next 12 months

In the next 12 months, how likely is it that you may become unemployed and not able to find a job within four weeks?

| Value | Category                              |
|:------|:--------------------------------------|
| 1     | Not at all likely                     |
| 2     | Not very likely                       |
| 3     | Somewhat likely                       |
| 4     | Very likely                           |
| 5     | Extremely likely                      |
| 6     | Never worked or currently not working |
| 9     | No answer\*                           |

\*) Missing value

```{r}
# Load packages
library(tidyverse)


# Load dataset 
data <- read_delim(
  "C:/Users/ADMIN/Desktop/DAT505/RR2/ESS.csv",
  delim = ";",
  escape_double = FALSE,
  trim_ws = TRUE
)
```

Now we will clean the data and keep only the variables relevent for our research question and hypothesis

```{r}

# 1. Convert relevant variables to numeric 
data <- data %>%
  mutate(across(
    c(w2gq12, w2gq8, w2gq1,
      w2eq15, w2eq16, w2eq17, w2eq18,
      hincfel, age, eduyrs),
    as.numeric
  ))

# 2. Recode missing values (CRONOS codes from the codebook)
data <- data %>%
  mutate(across(
    c(w2gq12, w2gq8, w2gq1,
      w2eq15:w2eq18,
      hincfel, age, eduyrs),
    ~ ifelse(. %in% c(77, 88, 99), NA, .)
  ))

# 3. Create key variables
data <- data %>%
  mutate(
    # Main outcome DV
    climate_efficacy = w2gq12,
    
    # Alternative outcome (for robustness maybe later)
    climate_fatalism = w2gq8,
    
    # Economic insecurity index
    econ_insecurity = rowMeans(select(., w2eq15:w2eq18), na.rm = TRUE),
    
    # Controls
    climate_worry = w2gq1,
    
    # Clean factors
    gender = factor(gndr),
    country = factor(cntry)
  )

# 4. Keep only relevant variables
data_clean <- data %>%
  select(
    climate_efficacy,
    climate_fatalism,
    econ_insecurity,
    hincfel,
    climate_worry,
    age,
    eduyrs,
    gender,
    country
  )

# 5. Dropping missing values (for regression)
data_clean <- data_clean %>%
  drop_na()

summary(data_clean)
view(data_clean)
```

We created an **economic insecurity index** by averaging respondents’ answers to the four hardship-risk questions (`w2eq15–w2eq18`), ignoring missing values, so higher scores reflect greater perceived financial insecurity.

Descriptive statistics

```{r}
# Create basic descriptive table
desc_table <- data_clean %>%
  summarise(
    climate_efficacy = mean(climate_efficacy, na.rm = TRUE),
    econ_insecurity  = mean(econ_insecurity, na.rm = TRUE),
    hincfel          = mean(hincfel, na.rm = TRUE),
    climate_worry    = mean(climate_worry, na.rm = TRUE),
    age              = mean(age, na.rm = TRUE),
    eduyrs           = mean(eduyrs, na.rm = TRUE)
  ) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "mean") %>%
  mutate(
    sd = c(
      sd(data_clean$climate_efficacy, na.rm = TRUE),
      sd(data_clean$econ_insecurity, na.rm = TRUE),
      sd(data_clean$hincfel, na.rm = TRUE),
      sd(data_clean$climate_worry, na.rm = TRUE),
      sd(data_clean$age, na.rm = TRUE),
      sd(data_clean$eduyrs, na.rm = TRUE)
    )
  )
desc_table
```

```{r}
library(knitr)
library(kableExtra)

# Clean + label table
desc_table_clean <- desc_table %>%
  mutate(
    Variable = recode(Variable,
      "climate_efficacy" = "Climate efficacy",
      "econ_insecurity"  = "Economic insecurity",
      "hincfel"          = "Income perception",
      "climate_worry"    = "Climate worry",
      "age"              = "Age",
      "eduyrs"           = "Years of education",
      "1"                = "Male",
      "2"                = "Female"
    )
  ) %>%
  # Order variables nicely
  mutate(Variable = factor(Variable, levels = c(
    "Climate efficacy",
    "Economic insecurity",
    "Income perception",
    "Climate worry",
    "Age",
    "Years of education",
    "Male",
    "Female"
  ))) %>%
  arrange(Variable)

# Display
kable(desc_table_clean, digits = 2,
      caption = "Descriptive Statistics (including demographics)") %>%
  kable_styling(full_width = FALSE, position = "left")
```

```{r}
# Create plot: climate efficacy
p1 = ggplot(data_clean, aes(x = climate_efficacy)) +
  geom_histogram(binwidth = 1,
                 fill = "darkred",
                 color = "black",
                 alpha = 0.85) +
  labs(title = "Distribution of Climate Efficacy",
       x = "Climate Efficacy (0–10)",
       y = "Number of Respondents") +
  theme_minimal()
p1
```

```{r}
# Create plot: economic insecurity
p2 = ggplot(data_clean, aes(x = econ_insecurity)) +
  geom_histogram(binwidth = 0.5,
                 fill = "darkgreen",
                 color = "black",
                 alpha = 0.85) +
  labs(title = "Distribution of Economic Insecurity",
       x = "Economic Insecurity Index",
       y = "Number of Respondents") +
  theme_minimal()
p2
```

```{r}
library(patchwork)

p_combined = (p1 | p2)
p_combined
```

```{r}
# Scatter plot with linear trend
p3 = ggplot(data_clean, aes(x = econ_insecurity, y = climate_efficacy)) +
  geom_point(alpha = 0.2,
             color = "darkblue",
             size = 1.5) +
  geom_smooth(method = "lm",
              color = "orange",
              fill = "black",
              alpha = 0.2,
              linewidth = 1.2) +
  labs(title = "Economic Insecurity and Climate Efficacy",
       subtitle = "Higher insecurity is associated with lower perceived agency",
       x = "Economic Insecurity",
       y = "Climate Efficacy") +
  theme_minimal()
p3

```
