---
title: "RR2 first draft"
format: 
  html:
    self_contained: true
    toc: true
    embed-resources: true
editor: visual
author: Chita Chaaban
date: Winter 2026
---

## First Draft:

### Introduction

In this report we examine **how economic insecurity shapes individuals’ perceived ability to act on climate change**. Rather than analyzing environmental attitudes in isolation, we try to focus on how material conditions may constrain individual agency. The CRONOS3 Wave 2 dataset is suited for this question, as it combines environmental attitudes from the “Make it Green” module with measures of economic insecurity from the “Make it Equal” module, allowing both dimensions to be analyzed jointly.

The main outcome variable is perceived climate efficacy, measured through respondents’ agreement with the statement that their own actions can help prevent worsening climate change (w2gq12), which could capture individual-level agency in the environmental domain. As an alternative measure, climate fatalism (w2gq8) captures the opposite perception, namely that it is too difficult for someone like oneself to contribute meaningfully.

The main explanatory/ independent variable is economic insecurity, constructed as an index combining four items capturing the perceived likelihood of experiencing financial hardship in the next twelve months (w2eq15–w2eq18), including difficulties in meeting basic needs, affording unexpected expenses, or maintaining employment. We also include subjective income perception (hincfel) to capture current economic evaluation, complementing the forward-looking insecurity index.

Climate change requires sustained individual and collective action, yet not everyone feels equally capable of contributing. While research has established that concern about climate change drives perceived efficacy, less attention has been paid to how material conditions shape this sense of agency. This matters because economic insecurity has risen sharply across Europe in recent years, and if financial precarity systematically undermines perceived climate efficacy, this has direct implications for how we understand climate disengagement.

Drawing on scarcity theory (Mullainathan & Shafir, 2020), we argue that individuals facing perceived financial hardship have reduced cognitive and motivational resources to engage with long-term collective problems. When meeting basic needs feels uncertain, climate action may feel not just irrelevant but genuinely out of reach.

This leads to our main hypothesis:

H1: Higher perceived economic insecurity is associated with lower climate self-efficacy and higher climate fatalism, even after controlling for climate concern.

The theoretical expectation is that economic insecurity reduces perceived agency: when individuals face uncertainty about their material conditions, their attention and resources are directed toward short-term constraints, which may limit their engagement with long-term collective issues such as climate change. To isolate this mechanism, the analysis controls for climate worry (w2gq1), ensuring that the estimated effect is not simply driven by differences in concern about climate change.

### Data and Preparation

The analysis uses the CRONOS3 Wave 2 (edition 1.3) dataset, a cross-national European survey conducted in 2024 across eleven countries. The dataset includes detailed measures of environmental attitudes, economic insecurity, and socio-demographic characteristics. The data was first restricted to the variables relevant to the research question. All selected variables were converted to numeric format where necessary, and missing values were recoded according to the survey coding scheme (77, 88, 99 treated as missing).

An index of economic insecurity was then constructed by taking the average of four items capturing perceived risks of financial hardship (w2eq15–w2eq18), ignoring missing values. This index is meant to provide us with a continuous measure of perceived economic vulnerability, where higher values indicate greater insecurity. The main outcome variable, climate efficacy, was directly taken from, while climate fatalism was retained as an alternative outcome for robustness checks. Control variables were also defined at this stage. Climate worry was also included as a control to distinguish concern from perceived agency. Then, additional socio-demographic variables, including age, gender, education, and country, were included to account for potential confounding factors.

Finally, the dataset was reduced to a clean analytical sample by keeping only the variables of interest and removing observations with missing values on these variables. This ensures consistency across all models and avoids variations in sample size between specifications.

```{r}
# Load packages
library(tidyverse)


# Load dataset 
data <- read_delim(
  "C:/Users/ADMIN/Desktop/DAT505/RR2/ESS.csv",
  delim = ";",
  escape_double = FALSE,
  trim_ws = TRUE
)
```

Now we will clean the data and keep only the variables relevent for our research question and hypothesis

```{r}

# 1. Convert relevant variables to numeric 
data <- data %>%
  mutate(across(
    c(w2gq12, w2gq8, w2gq1,
      w2eq15, w2eq16, w2eq17, w2eq18,
      hincfel, age, eduyrs),
    as.numeric
  ))

# 2. Recode missing values (CRONOS codes from the codebook)
data <- data %>%
  mutate(across(
    c(w2gq12, w2gq8, w2gq1,
      w2eq15:w2eq18,
      hincfel, age, eduyrs),
    ~ ifelse(. %in% c(77, 88, 99), NA, .)
  ))

# 3. Create key variables
data <- data %>%
  mutate(
    # Main outcome DV
    climate_efficacy = w2gq12,
    
    # Alternative outcome (for robustness maybe later)
    climate_fatalism = w2gq8,
    
    # Economic insecurity index
    econ_insecurity = rowMeans(select(., w2eq15:w2eq18), na.rm = TRUE),
    

    # Controls
    climate_worry = w2gq1,
    
    # Clean factors
    gender = factor(gndr),
    country = factor(cntry),
    Income_perception = factor(hincfel)
  )

# 4. Keep only relevant variables
data_clean <- data %>%
  select(
    climate_efficacy,
    climate_fatalism,
    econ_insecurity,
    Income_perception,
    climate_worry,
    age,
    eduyrs,
    gender,
    country
  )

# 5. Dropping missing values (for regression)
data_clean <- data_clean %>%
  drop_na()

summary(data_clean)
view(data_clean)
```

We created an **economic insecurity index** by averaging respondents’ answers to the four hardship-risk questions (`w2eq15–w2eq18`), ignoring missing values, so higher scores reflect greater perceived financial insecurity.

```{r}
 # Cronbach's Alpha for the economic insecurity index
  
library(psych)

econ_items <- data %>%
  select(w2eq15, w2eq16, w2eq17, w2eq18)

alpha_econ <- psych::alpha(econ_items)

alpha_econ$total$raw_alpha
```

To assess whether the four items used to construct the economic insecurity index formed a coherent scale, I tested their internal consistency using Cronbach’s alpha. The four-item scale showed acceptable internal consistency (`α = 0.78`), supporting the decision to combine these items into a single index.

Descriptive statistics

```{r}
# Create basic descriptive table
desc_table <- data_clean %>%
  summarise(
    climate_efficacy  = mean(climate_efficacy, na.rm = TRUE),
    econ_insecurity   = mean(econ_insecurity, na.rm = TRUE),
    Income_perception = mean(Income_perception, na.rm = TRUE),
    climate_worry     = mean(climate_worry, na.rm = TRUE),
    age               = mean(age, na.rm = TRUE),
    eduyrs            = mean(eduyrs, na.rm = TRUE)
  ) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "mean") %>%
  mutate(
    sd = c(
      sd(data_clean$climate_efficacy, na.rm = TRUE),
      sd(data_clean$econ_insecurity, na.rm = TRUE),
      sd(data_clean$hincfel, na.rm = TRUE),
      sd(data_clean$climate_worry, na.rm = TRUE),
      sd(data_clean$age, na.rm = TRUE),
      sd(data_clean$eduyrs, na.rm = TRUE)
    )
  )
desc_table
```

```{r}
# Create plot: climate efficacy
p1 = ggplot(data_clean, aes(x = climate_efficacy)) +
  geom_histogram(binwidth = 1,
                 fill = "darkred",
                 color = "black",
                 alpha = 0.85) +
  labs(title = "Distribution of Climate Efficacy",
       x = "Climate Efficacy (0–10)",
       y = "Number of Respondents") +
  theme_minimal()
p1
```

```{r}
# Create plot: economic insecurity
p2 = ggplot(data_clean, aes(x = econ_insecurity)) +
  geom_histogram(binwidth = 0.5,
                 fill = "darkgreen",
                 color = "black",
                 alpha = 0.85) +
  labs(title = "Distribution of Economic Insecurity",
       x = "Economic Insecurity Index",
       y = "Number of Respondents") +
  theme_minimal()
p2
```

```{r}
library(patchwork)

p_combined = (p1 | p2)
p_combined
```

Interpretation

Climate efficacy appears moderately centered, with most respondents clustering around mid-to-high values, suggesting that individuals generally report some level of perceived agency in addressing climate change. The distribution is relatively balanced, though slightly concentrated around the middle of the scale.

In contrast, economic insecurity is more skewed toward lower values, indicating that most respondents report relatively low to moderate levels of perceived financial risk, with fewer individuals experiencing high insecurity. The right tail suggests that a smaller group faces substantially higher levels of economic vulnerability.

Taken together, this suggests that while perceived climate agency is relatively widespread, economic insecurity is more unevenly distributed across respondents. This is indicating that most respondents are at low to moderate levels in terms of economic insecurity.

```{r}
# Scatter plot with linear trend
p3 = ggplot(data_clean, aes(x = econ_insecurity, y = climate_efficacy)) +
  geom_point(alpha = 0.2,
             color = "darkblue",
             size = 1.5) +
  geom_smooth(method = "lm",
              color = "orange",
              fill = "black",
              alpha = 0.2,
              linewidth = 1.2) +
  labs(title = "Economic Insecurity and Climate Efficacy",
       subtitle = "Higher insecurity is associated with lower perceived agency",
       x = "Economic Insecurity",
       y = "Climate Efficacy") +
  theme_minimal()
p3

```

Interpretation:

The scatter plot shows a weak positive relationship between economic insecurity and climate efficacy, as indicated by the upward-sloping regression line. This suggests that, at the bivariate level, higher economic insecurity is not associated with lower perceived agency, and if anything, the relationship appears slightly positive.

However, the relationship is relatively flat and the data are widely dispersed, this means that substantial variation in climate efficacy at all levels of insecurity. This suggests that economic insecurity alone is not a strong predictor of perceived climate agency.

In relation to the hypothesis, these descriptive results do not provide clear support for the expected negative relationship. Instead, they highlight the need to account for additional factors, particularly climate worry, to better isolate the effect of economic insecurity on perceived agency.

### **Inferential stats:**

Because the gender and income perception have 9 coded as NA, I changed it (probably should move it to the data cleaning section).

```{r}
data_clean <- data_clean %>%
  mutate(
    # Convert to numeric FIRST
    gender = as.numeric(as.character(gender)),
    Income_perception = as.numeric(as.character(Income_perception)),
    
    # Remove invalid codes
    gender = na_if(gender, 9),
    Income_perception = na_if(Income_perception, 9),
    
    # Convert to factor AFTER cleaning
    gender = factor(gender),
    Income_perception = factor(Income_perception)
  ) %>%
  drop_na(gender, Income_perception)
```

```{r}
# Model 1: Bivariate relationship 
model1 <- lm(climate_efficacy ~ econ_insecurity, data = data_clean)

summary(model1)
```

In Model 1, the bivariate relationship between economic insecurity and climate efficacy is small and not statistically significant (*B* = 0.075, p = 0.304). This indicates that, on its own, economic insecurity is not associated with lower perceived climate agency, and the direction of the relationship is slightly positive rather than negative.

```{r}
# Model 2: Add climate worry (core control)
model2 <- lm(climate_efficacy ~ econ_insecurity + climate_worry, data = data_clean)

summary(model2)
```

In Model 2, after controlling for climate worry, the coefficient for economic insecurity remains positive and non-significant, while climate worry emerges as a strong and statistically significant predictor (*B*= 0.489, p \< 0.001). This suggests that perceived agency is more closely related to concern about climate change than to economic insecurity.

```{r}
# Model 3: Full model with controls
model3 <- lm(climate_efficacy ~ econ_insecurity + Income_perception + climate_worry +
               age + eduyrs + gender + country,
             data = data_clean)
summary(model3)
```

In the full model (Model 3), the effect of economic insecurity remains small and non-significant (*B* = 0.075, p = 0.304), providing no support for the hypothesis that higher insecurity reduces perceived climate efficacy. In contrast, climate worry continues to have a strong positive association with efficacy, and age shows a small but significant negative effect (*B* = -0.007, p \< 0.01), indicating that older individuals report slightly lower perceived agency. Gender is also significant, with women reporting higher climate efficacy than men (*B* = 0.478, p \< 0.01).

Overall, these results do not support the main hypothesis. Economic insecurity does not appear to reduce perceived climate agency, and the expected negative relationship is not observed. Instead, perceived efficacy is more strongly explained by climate concern than by economic conditions.

### Robustness check:

```{r}
model4 <- lm(climate_fatalism ~ econ_insecurity + climate_worry +
               age + eduyrs + gender + country,
             data = data_clean)
summary(model4)

ggplot(data_clean, aes(x = econ_insecurity, y = climate_efficacy)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm")

```

The robustness check confirms the main findings. The relationship between economic insecurity and climate efficacy remains weak and slightly positive, as indicated by the upward-sloping regression line. This is consistent with the regression results, where the coefficient for economic insecurity was small and not statistically significant (*B* = 0.075, p = 0.304).

The dispersion of the data remains high across all levels of insecurity, suggesting that individuals with both low and high economic insecurity report similar levels of climate efficacy. This reinforces the conclusion that economic insecurity is not a strong predictor of perceived climate agency.

The robustness check supports the initial results: the expected negative relationship is not observed, and the findings remain stable across specifications.

### **Classification: Predictive modelling** 

**Train-test split for the main model:**

To see whether the main model from the analysis can predict new observations, not just fit the sample used to estimate it, we first split the data into a training set and a test set. Then we estimate the model on the training set, and then evaluate how well it predicts climate efficacy in the test set. Since the outcome is continuous, I use RMSE and R-squared to assess predictive performance.

```{r}
# Load the packages used for this section
library(caret)
library(knitr)

# Set a seed so the split is reproducible
set.seed(1234)

# Create an 80/20 train-test split
train_index <- createDataPartition(data_clean$climate_efficacy, p = 0.8, list = FALSE)

train_data <- data_clean[train_index, ]
test_data  <- data_clean[-train_index, ]

# Fit the main linear regression model on the training set
main_model_train <- lm(
  climate_efficacy ~ econ_insecurity + Income_perception + climate_worry +
    age + eduyrs + gender + country,
  data = train_data
)

# Predict climate efficacy in the test set
test_pred <- predict(main_model_train, newdata = test_data)

# Compute predictive performance
rmse_test <- RMSE(test_pred, test_data$climate_efficacy)
r2_test   <- R2(test_pred, test_data$climate_efficacy)

# Put the results into a small table
test_results <- data.frame(
  RMSE = rmse_test,
  R_squared = r2_test
)

# Display the table
kable(
  test_results,
  digits = 3,
  caption = "Predictive performance on the test set"
)
```

The predictive performance of the model on the test set is relatively weak. The RMSE of 2.57 indicates that, on average, the model’s predictions deviate from the true climate efficacy scores by about 2.6 points. Given that the outcome is measured on a 0–10 scale, this represents a fairly large prediction error.

The R-squared of 0.03 shows that the model explains only 3% of the variance in climate efficacy in unseen data. This is very low, indicating that the model has limited ability to capture the factors that drive differences in climate efficacy across individuals.

These results suggest that while the model may still be useful for identifying associations between variables (as in the inferential analysis), it performs poorly as a predictive tool. In other words, the variables included in the model, such as economic insecurity and climate worry, do not provide strong predictive power for accurately forecasting individual levels of climate efficacy.

**Cross-validated predictive performance of the main model:**

While the train-test split provides a first indication of predictive performance, it relies on a single random partition of the data. To obtain a more reliable estimate, I next use 10-fold cross-validation. This approach repeatedly divides the dataset into training and validation subsets, allowing the model to be evaluated across multiple data partitions rather than a single split.

```{r}
# Define cross-validation settings
train_control_cv <- trainControl(
  method = "cv",
  number = 10
)

# Train the model using 10-fold cross-validation
cv_model <- train(
  climate_efficacy ~ econ_insecurity + Income_perception + climate_worry +
    age + eduyrs + gender + country,
  data = data_clean,
  method = "lm",
  metric = "RMSE",
  trControl = train_control_cv
)

print(cv_model)
```

The cross-validation results confirm that the predictive performance of the model remains limited. The RMSE of 2.45 indicates that, on average, the model’s predictions deviate from the true values of climate efficacy by approximately 2.5 points on the 0–10 scale. This is a relatively large prediction error given the range of the outcome variable.

The R-squared is approximately 0.096, meaning that the model explains about 9.6% of the variance in climate efficacy across the different validation folds. While this is slightly higher than the test set estimate, it still indicates that the vast majority of the variation in climate efficacy remains unexplained.

The MAE of 1.99 provides a complementary measure of prediction error and suggests that the average absolute deviation between predicted and observed values is around 2 points. This reinforces the conclusion that prediction errors are substantial in magnitude.

Overall, these cross-validation results are consistent with the earlier train-test split findings. They suggest that the model has weak predictive performance, even when evaluated across multiple data partitions. This indicates that the limited predictive power is not due to a particular data split, but rather reflects a more general limitation of the model and the available predictors.

**Bootstrap validation**

To further assess the robustness of these findings, I implement bootstrap validation. Unlike cross-validation, which partitions the data into fixed folds, bootstrap validation repeatedly resamples the dataset with replacement, providing an alternative estimate of predictive performance.

```{r}
#Define bootstrap settings

train_control_boot <- trainControl( method = "boot", number = 100 )

#Train the model using bootstrap validation

boot_model <- train( climate_efficacy ~ econ_insecurity + Income_perception + climate_worry + age + eduyrs + gender + country, data = data_clean, method = "lm", metric = "RMSE", trControl = train_control_boot )

print(boot_model)
```

The bootstrap validation results further confirm the limited predictive performance of the model. The RMSE is 2.47, which is very close to the value obtained through cross-validation. This indicates that the average prediction error remains substantial, with predicted values deviating from observed climate efficacy by roughly 2.5 points.

The R-squared is approximately 0.071, meaning that the model explains only about 7.1% of the variance in climate efficacy under bootstrap resampling. This is slightly lower than the estimate obtained through cross-validation, suggesting that the model’s predictive capacity is not only weak but also somewhat sensitive to the validation method used.

The MAE of 2.00 reinforces this conclusion, indicating that the average absolute prediction error remains around 2 points, which is considerable given the scale of the outcome variable.

The consistency of RMSE and MAE across validation methods, combined with persistently low R-squared values, suggests that the model’s limited predictive performance is robust. These results indicate that the predictors included in the model capture only a small portion of the variation in climate efficacy, regardless of the resampling strategy employed.

**A data-driven approach: lasso regression:**

The previous analyses evaluated the predictive performance of a theory-driven model. Noow we extend the analysis by adopting a more data-driven approach. Specifically, we use lasso regression, which applies a penalty to the model coefficients and shrinks less informative predictors toward zero. This allows the model to identify a more parsimonious specification that may improve predictive performance.

```{r}
library(glmnet)

# Create predictor matrix
x <- model.matrix(
  climate_efficacy ~ econ_insecurity + Income_perception + climate_worry +
    age + eduyrs + gender + country,
  data = data_clean
)[, -1]

y <- data_clean$climate_efficacy

set.seed(1234)

# Cross-validated lasso
cv_lasso <- cv.glmnet(
  x, y,
  alpha = 1,
  family = "gaussian",
  nfolds = 10
)

# Lambda values
data.frame(
  Lambda_min = cv_lasso$lambda.min,
  Lambda_1se = cv_lasso$lambda.1se
)

# Coefficients
coef(cv_lasso, s = "lambda.min")
```

Interpretation of lasso results (lambda): The optimal penalty parameter selected through cross-validation is λ = 0.0436 (lambda_min), which corresponds to the model with the lowest prediction error. The alternative, λ = 0.2802 (lambda_1se), represents a more conservative specification with stronger penalization. The relatively low value of λ at lambda.min suggests that only moderate shrinkage is needed to optimize predictive performance.

Interpretation of lasso coefficients: The lasso results show that several predictors are shrunk to zero, meaning they do not contribute meaningfully to prediction once the penalty is applied. In particular, most categories of Income_perception are eliminated, suggesting that this variable adds little predictive value beyond other controls.

Among the retained predictors, climate_worry remains the strongest positive predictor, confirming its central role in explaining variation in climate efficacy. gender and several country fixed effects also remain important, indicating that demographic and contextual differences contribute to prediction.

In contrast, econ_insecurity is retained but with a very small coefficient, suggesting that it has only a limited contribution to predictive performance. Similarly, age and education have very small effects, indicating weak predictive relevance.

**Predictive performance of the lasso model:**

To directly compare the predictive performance of the lasso model with the original specification, we evaluate it using the same train-test approach.

```{r}

# Set a seed
set.seed(1234)

# Randomly split the data into 80% training data and 20% test data
train_index <- createDataPartition(data_clean$climate_efficacy, p = 0.8, list = FALSE)

# Create the training and test datasets
train_data <- data_clean[train_index, ]
test_data  <- data_clean[-train_index, ]

# Build the predictor matrix for the training data
# model.matrix() turns categorical predictors into dummy variables
# We remove the first column because it is the intercept
x_train <- model.matrix(
  climate_efficacy ~ econ_insecurity + Income_perception + climate_worry +
    age + eduyrs + gender + country,
  data = train_data
)[, -1]

# Define the outcome variable for the training data
y_train <- train_data$climate_efficacy

# Build the predictor matrix for the test data using the same predictors
x_test <- model.matrix(
  climate_efficacy ~ econ_insecurity + Income_perception + climate_worry +
    age + eduyrs + gender + country,
  data = test_data
)[, -1]

# Define the outcome variable for the test data
y_test <- test_data$climate_efficacy

# Fit a lasso regression on the training data
# alpha = 1 specifies lasso
# family = "gaussian" is used because the outcome is continuous
# nfolds = 10 means that lambda is selected through 10-fold cross-validation
cv_lasso_train <- cv.glmnet(
  x_train, y_train,
  alpha = 1,
  family = "gaussian",
  nfolds = 10
)

# Use the fitted lasso model to predict climate efficacy in the test set
# lambda.min gives the value of lambda with the lowest cross-validated error
lasso_pred <- predict(cv_lasso_train, newx = x_test, s = "lambda.min")

# Calculate RMSE to measure the average prediction error
rmse_lasso <- RMSE(as.numeric(lasso_pred), y_test)

# Calculate R-squared to measure how much variance is explained in the test set
r2_lasso   <- R2(as.numeric(lasso_pred), y_test)

# Display the predictive performance of the lasso model
data.frame(
  RMSE = rmse_lasso,
  R_squared = r2_lasso
)
```

The predictive performance of the lasso model remains very weak. The RMSE of 2.56 indicates that the average prediction error is virtually identical to that of the main model, with predicted values deviating from observed climate efficacy by about 2.5 points. The R-squared is 0.031, meaning that the model explains only around 3.1% of the variance in climate efficacy in the test set. This is nearly identical to the performance of the original model, suggesting that the lasso specification does not improve predictive accuracy.

**Conclusion of Predictive modelling:**

A better predictive model would not necessarily involve a different estimation technique, but rather the inclusion of more theoretically relevant predictors that capture the psychological and behavioral dimensions of climate efficacy. The current results suggest that economic and demographic variables alone are insufficient to accurately predict individual-level variation in climate efficacy.
